Effects of the body force on the pedestrian and evacuation dynamics

### 2020-02-07 paper.tex
reemplazo de compression por overlap
agregue las correcciones de guille del 6/2

The presence of the FIF phenomenon for only ``soft'' individuals opens many questions 
on the microscopic dynamics of pedestrians. We may presume that contacts 
between pedestrians are quite different for soft individuals than for stiff 
individuals. Thus, we proceed to study the dynamics of contacting pedestrians, regardless of the compression 
effects. We will assimilate the pedestrians as nodes and the whole crowd as a 
network. We will link any two individuals whenever they get in physical 
contact (\textit{i.e.} $r_{ij} \leq R_{ij}$).\\

@@

We may summarize this Section as follows. We explored the parameter space along $v_d$ and $k_n$ for the bottleneck situation. This is similar to explore the unitless space $(\mathcal{A},\mathcal{K}_c)$. Soft pedestrians attain a FIS or FIS+FIF evacuation pattern but, stiff pedestrians (for any fixed $\mathcal{A}$ value within the explored range) exhibit a single FIF pattern. Stiffness affects either the presence of the blocking clusters and the pedestrians overlap. The less they overlap, the less intense becomes the sliding friction among them. Additionally, the fewer the blocking clusters the easier they get out. \\

@@

We present in this section the results corresponding to the corridor geometry. We show the effects of modifying the body force coefficient $k_n$ on the collective dynamics.\\

We computed the contact network in the same way as in the bottleneck geometry. Fig.~\ref{degree_dens} shows the mean degree as a function of the global density for different $k_n$ values (see caption for details). The degree is zero when the pedestrians do not touch each other. When the density reaches 4.5, a few pedestrians start to touch each other increasing the mean degree. As the density continuous increasing, the mean degree increases towards the asymptotic degree six. Degree six corresponds to the maximum packing density for identical hard circles.\\

Fig. \ref{degree_dens} exhibits a somewhat opposite behavior with respect to the bottleneck geometry, for a given value of density. Here, the stiffer the pedestrians the higher the degree. In the bottleneck geometry, however, the stiffer the pedestrians, the lower the mean degree. Besides, Fig. \ref{overlap_dens} shows that the mean overlap increases as the global density increases. This behavior appears similar to the bottleneck situation, in the sense that the higher the value of $k_n$ the less they overlap. The same reasoning in Section \ref{bottleneck} apply to this situation.  \\

\begin{figure}[!htbp]
\centering
    \subfloat[]{\includegraphics[width=0.49\columnwidth]{./degree_vs_dens_multi_kn.eps}\label{degree_dens}}\ 
    \subfloat[]{\includegraphics[width=0.49\columnwidth]{./overlap_vs_dens_multi_kn.eps}\label{overlap_dens}}\\
\caption[width=0.47\columnwidth]{(a) Mean degree as a function of the global density for different $k_n$ values. (b) mean overlap as a function of the global density. The global density is the total number of pedestrians per unit area. The mean values are averages over all the pedestrians and over time once the system reached the stationary state. Both measurements correspond to corridor geometry with desired velocity $v_d=1$. Color online only.}
\label{network_corridor}
\end{figure}

In order to get a better insight on how the pedestrians contact to each other, we present in Fig.~\ref{network_corridor} two snapshots of the corridor at the stationary situation. Fig.~\ref{network_d6_knE5} corresponds to $k_n=1.2\,$E5. It can be seen that increasing the stiffness increases the number of contacts between pedestrians (say, the network connectivity). The number of nodes attaining degree 6 is far larger in the case of $k_n=1.2\,$E5 than the case of $k_n=0$. \\

This result is exactly the opposite of what we get in the bottleneck geometry. This phenomenon occurs because, in the corridor, the lateral walls act like a containment barrier that forces the pedestrian to increase their contacts. Whereas in the bottleneck there are no lateral walls, thus, when increasing the stiffness they have more space to detach from each other, reducing the connectedness (hence reducing the mean degree). \\

\begin{figure}[!htbp]
\centering
    \subfloat[]{\includegraphics[width=0.49\columnwidth]{./network_d6_kn0.eps}\label{network_d6_kn0}}\ 
    \subfloat[]{\includegraphics[width=0.49\columnwidth]{./network_d6_knE5.eps}\label{network_d6_knE5}}\\
\caption[width=0.47\columnwidth]{Contact network of the pedestrians along the corridor at time $t=50\,$s. The global density was $\rho=6$.  The lines that connect the nodes (pedestrians) represent the contact between them. The colors stand for the degree of the node (the number of pedestrians that are in contact with him/her). The corridor was 28$\,$m $\times$ 22$\,$m with periodic boundary conditions and $v_d=1$. (a) corresponds to a simulation without body force and (b) corresponds to a simulation with $k_n=1.2\,$E5. The friction coefficient and the other SFM parameters are the same as in Section \ref{bottleneck}. Color online only.}
\label{network_corridor}
\end{figure}

In the corridor geometry, the crowd gets more connected if the pedestrians become stiffer. We studied the network properties of the contact graph arising from the positions $(x,y)$ of every pedestrian. We found that a good descriptor to exhibit the connectivity is the number of triangles per node. A triangle is defined as a structure composed of three nodes, all in contact with each other. Fig.~\ref{triangles} shows the triangles per node as a function of the global density for different stiffness values $k_n$. This descriptor allows distinguishing states with the same global density but different stiffness levels. In other words, it is a descriptor very sensitive to changes in stiffness that allow expressing the connectivity of the system even better than the mean degree. \\

This result is consistent with the results obtained in Ref.~\cite{pugnaloni_2013} where the authors explore several topological and geometrical descriptors capable of properly describing granular systems with different stiffness levels. While our study system is self-propelled, there are analogies with the system studied in Ref.~\cite{pugnaloni_2013}. Like the authors, we recommend using topological descriptors (in particular triangles per node) to describe systems with equal global density and different stiffness levels.\\

\begin{figure}[htbp!]
\centering
\includegraphics[width=0.7\columnwidth]
{./triangles.eps}
\caption{\label{triangles} Triangles per node as a function of the global density. A triangle in a network is defined as three nodes all connected. Each marker corresponds to a different stiffness $k_n$ (see the label in the plot). The measurements correspond to a corridor of 28$\,$m $\times$ 22$\,$m with periodic boundary conditions and $v_d=1$. Color online only. }
\end{figure}

The most important result is that increasing the $k_n$ value reduces the flow (hence reduces the walking speed of the overall crowd) in the corridor geometry. Fig.~\ref{kn_vs_vx_corridor} shows the mean velocity in the direction of displacement $v_x$ as a function of the stiffness for different global density levels. $v_x$ reduces as $k_n$ increases as opposed to what happens in the bottleneck geometry. There is little to no change in the speed from zero to $k_n=2.6\,$E4, from this value on, the diminishing of velocity is more abrupt.\\

The reason why the velocity reduces as the stiffness of the pedestrians' increases is that it becomes harder for pedestrians to walk forward. The difficulty arises because the pedestrian feels a much higher repulsion from the pedestrian ahead (due to the body force increment). Thus, they have no choice but to move at (almost) the same speed as the individuals they have at their sides. This microscopic mechanism explains the macroscopic behavior that turns the crowd from a ``fluid-like state'' into a ``solid-like state''. When the stiffness is high enough (say $k_n=1.2\,$E6), all the pedestrians walk at the same velocity. The pedestrians that walk in physical contact with the wall are the ones who determine the velocity of the whole crowd.\\

\begin{figure}[htbp!]
\centering
\includegraphics[width=0.7\columnwidth]{./kn_vs_vx_corridor.eps}
\caption{\label{kn_vs_vx_corridor} Mean velocity in the longitudinal coordinate $(v_x)$ as a function of the stiffness $k_n$ for three different global densities (see label in the plot). The measurements correspond to a corridor of 28$\,$m $\times$ 22$\,$m with periodic boundary conditions and $v_d=1$. Color online only.      }
\end{figure}

>>>>

The presence of the FIF phenomenon for only ``soft'' individuals opens many questions 
on the microscopic dynamics of pedestrians. We may presume that contacts 
between pedestrians are quite different for soft individuals than for stiff 
individuals. Thus, we proceed to study the dynamics of contacting pedestrians, regardless of the overlap 
effects. We will assimilate the pedestrians as nodes and the whole crowd as a 
network. We will link any two individuals whenever they get in physical 
contact (\textit{i.e.} $r_{ij} \leq R_{ij}$).\\

@@

We may summarize this Section as follows. We explored the parameter space along $v_d$ and $k_n$ for the bottleneck situation. This is similar to explore the unitless plane $(\mathcal{A},\mathcal{K}_c)$. Soft pedestrians attain a FIS or FIS+FIF evacuation pattern but, stiff pedestrians (for any fixed $\mathcal{A}$ value within the explored range) exhibit a single FIF pattern. Stiffness affects either the presence of the blocking clusters and the pedestrians overlap. The less they overlap, the less intense becomes the sliding friction among them. Additionally, the fewer the blocking clusters the easier they get out. \\

@@

We present in this section the results corresponding to the corridor geometry. 
We show the effects of modifying the body force coefficient $k_n$ on the 
collective dynamics. We will keep $v_d$ (or $\mathcal{A}$) 
fixed along this Section. \\

We first computed the contact network in the same way 
as in the bottleneck geometry. Fig.~\ref{degree_dens} shows the 
mean degree as a function of the global density for different $k_n$ values (see 
caption for details). The mean degree vanishes at very low 
densities because the pedestrians do not touch each other. When the density 
surpasses 4.5, a few pedestrians start to touch each other, 
raising the mean degree. As the density continuous increasing, 
the mean degree approaches the asymptotic value 
of six. Degree six corresponds to the maximum packing density for identical 
hard circles.\\

\subfloat[]{\includegraphics[width=0.49\columnwidth]
{./degree_vs_dens_multi_kn.eps}\label{degree_dens}}\ 
\subfloat[]{\includegraphics[width=0.49\columnwidth]
{./triangles.eps}\label{triangles}}\\
% \subfloat[]{\includegraphics[width=0.49\columnwidth]
%{./overlap_vs_dens_multi_kn.eps}\label{overlap_dens}}\\
\caption[width=0.47\columnwidth]{(a) Mean degree as a function 
of the global density for different $k_n$ values. (b)  Triangles per node as a 
function of the global density. The global density is the total number of 
pedestrians per unit area. The mean values are averages over all the pedestrians 
and over time once the system reached the stationary state. The measurements 
correspond to a corridor of 28~m $\times$ 22~m with periodic boundary conditions 
and $v_d=1$. Color online only.}
\label{network_corridor}
\end{figure}

\textcolor{red}{It is worth noting} from Fig.~\ref{degree_dens} that ``stiff''
pedestrians exhibit a sharper transition to the maximum packing density than 
the ``soft'' pedestrians. We further checked this phenomenon by counting the 
number of loops of three nodes (say, the triangles) present in the network. 
This magnitude was reported to be a sensitive magnitude for characterizing the 
jamming transition in a compressed granular packing (see 
Ref.~\cite{pugnaloni_2013} for details).  As can be seen in 
Fig.~\ref{triangles}, the ``stiff'' pedestrians share more triangles 
than the ``soft'' ones.\textcolor{red}{ The maximum number of allowed triangles is 6, 
corresponding to a hexagonal arrangement.}    \\

In order to get a better insight on how the pedestrians contact to each other, 
we present in Fig.~\ref{network_corridor} two snapshots of the corridor at the 
stationary situation. Fig.~\ref{network_d6_kn0} corresponds to 
$k_n=0$, while Fig.~\ref{network_d6_knE5} corresponds to $k_n=1.2$~E5. The 
former shows a somewhat disordered network, while the latter exhibits an almost 
completely ordered lattice. The missing triangles in Fig.~\ref{network_d6_kn0} 
are replaced by other polygons of more than three edges.    \\

\begin{figure}[!htbp]
\centering
    \subfloat[]{\includegraphics[width=0.49\columnwidth]{./network_d6_kn0.eps}\label{network_d6_kn0}}\ 
    \subfloat[]{\includegraphics[width=0.49\columnwidth]{./network_d6_knE5.eps}\label{network_d6_knE5}}\\
\caption[width=0.47\columnwidth]{Contact network of the pedestrians along the corridor at time $t=50\,$s. The global density was $\rho=6$.  The lines that connect the nodes (pedestrians) represent the contact between them. The colors stand for the degree of the node (the number of pedestrians that are in contact with him/her). The corridor was 28~m $\times$ 22~m with periodic boundary conditions and $v_d=1$. (a) corresponds to a simulation without body force and (b) corresponds to a simulation with $k_n=1.2$~E5. The friction coefficient and the other SFM parameters are the same as in Section \ref{bottleneck}. Color online only.}
\label{network_corridor}
\end{figure}

We find these topological magnitudes useful for comparing the 
corridor geometry with respect to the bottleneck geometry. A re-examination of 
Fig.~\ref{degree_dens} (corridor) and Fig.~\ref{degree_vd} (bottleneck) reveal 
that the pedestrian stiffness $k_n$ affects differently the way they contact 
each other. The mean degree increases for ``stiff'' pedestrians moving along 
the corridor (at densities above $5\,$p/m$^2$). Conversely, the mean degree 
increases for ``soft'' pedestrians in the bottleneck situation.\\

Fig.~\ref{network_bottleneck} and 
Fig.~\ref{network_corridor} illustrate the connectivity differences between 
the bottleneck and the corridor situation. The bulk in 
Fig.~\ref{network_bottleneck} appears more heavily connected among ``soft'' 
pedestrians than among ``stiff'' pedestrians (see both snapshots at $v_d=10$). 
The opposite occurs in Fig.~\ref{network_corridor}. This discrepancy seems to 
be related to the boundary conditions, since the same SFM parameters were 
applied on both situations. We may speculate that this phenomenon occurs 
because, in the corridor, the lateral walls act like a containment barrier that 
forces the ``stiff'' pedestrian to increase his (her) contacts. On the contrary, 
no real containment walls exist in the bottleneck situation (regardless the side 
walls). \\

We can test our hypothesis by computing the mean overlap. 
Fig.~\ref{overlap_dens} shows this magnitude for the corridor situation. Notice 
that ``soft'' pedestrians attain a more intruding overlap than the ``stiff'' 
ones, as expected. This is in agreement with Fig.~\ref{overlap_vd} for the 
bottleneck situation (at a fixed value of $v_d$). The curves in 
Fig.~\ref{overlap_vd}, however, do not meet each other as in  
Fig.~\ref{overlap_dens}. The curve meeting attains for space limitations in 
the corridor, since overlapping occurs (almost) independently of the value of
$k_n$. This is clearly not the case in the bottleneck situation. \\

\begin{figure}[htbp!]
\centering
\includegraphics[width=0.7\columnwidth]
{./overlap_vs_dens_multi_kn.eps}
\caption{\label{overlap_dens} Mean overlap as a function of the 
global density. The global density is the total number of pedestrians per unit 
area. The mean values are averages over all the pedestrians and over time once 
the system reached the stationary state. Both measurements correspond to 
the corridor geometry with desired velocity $v_d=1$. Color 
online only. }
\end{figure}

The above results correspond to a milestone in the 
investigation. The connectivity differences between the bottleneck geometry 
and the corridor geometry were not expected and opens two major questions: how 
does the body stiffness interact with the walls, and consequently, how does 
this affect the flux across the corridor.\\

We start by computing the mean velocity across the corridor.  
Fig.~\ref{kn_vs_vx_corridor} shows the mean velocity 
$\langle v_x\rangle$ (parallel to the corridor) as a function 
of the stiffness for different global density levels. This 
plot shows a flat pattern for $k_n<10^4$ and a slowing down above this 
threshold. Notice from Fig.~\ref{kn_vs_vx_bottleneck} that this is opposed
to what happens in the bottleneck geometry. \\

\begin{figure}[htbp!]
\centering
\includegraphics[width=0.7\columnwidth]{./kn_vs_vx_corridor.eps}
\caption{\label{kn_vs_vx_corridor} Mean velocity in the longitudinal coordinate $(v_x)$ as a function of the stiffness $k_n$ for three different global densities (see label in the plot). The measurements correspond to a corridor of 28~m $\times$ 22~m with periodic boundary conditions and $v_d=1$. Color online only.      }
\end{figure}

The $\langle v_x\rangle$ values in Fig.~\ref{kn_vs_vx_corridor} 
were computed at densities $\rho\geq 5.5$. According to 
Fig.~\ref{overlap_dens}, these densities attain relevant overlaps. Furthermore, 
stiffness values $k_n\geq 10^4$ move the system to a more heavily connected 
stage, and thus, pedestrians have no choice but to walk at 
(almost) the same speed as his (her) neighbors. This 
behavior may be envisaged as the passage from a released 
walking movement to a tigh movement as the stiffness increases. A more 
physical picture would assimilate te former as a ``fluid-like 
state'' and the latter as a ``solid-like state''. When the stiffness is 
very high (say $k_n=$1.2~E6), all the pedestrians 
are expected to walk at a common velocity. The 
pedestrians that walk in physical contact with the wall are the ones who 
determine the velocity of the whole crowd, as discussed 
below.\\
